\pkg{partykit}: A Modular Toolkit for Recursive Partytioning in \proglang{R}

### 2014-03-26 pkg/partykit/inst/mloss/mloss2.Rnw
first draft complete
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@555 255f4249-2263-4417-b1ea-8498779e27a7

\begin{keywords}
open-source software, recursive partitioning, regression trees, classification trees, decision trees, statistical learning. 
\end{keywords}

@@

In the more than fifty years since \cite{Morgan+Sonquist:1963}
published their seminal paper on ``automatic interaction detection'',
a wide range of methods has been suggested that is usually
termed ``recursive partitioning'' or ``decision trees'' or
``tree(-structured) models'' etc. Particularly influential were the
algorithms CART \citep[classification and regression trees,][]{Breiman+Friedman+Olshen:1984},
C4.5 \citep{Quinlan:1993}, QUEST/GUIDE \citep{Loh+Shih:1997,Loh:2002},
and CTree \citep{Hothorn+Hornik+Zeileis:2006} among many others
\citep[see][for a recent overview]{Loh:2014}. Reflecting the heterogeneity
of conceptual algorithms, a wide range of computational implementations
in various software systems emerged: Typically the original authors of an
algorithm also provide accompanying software but many software systems,
including \pkg{Weka} \citep{Witten+Frank:2005} or \proglang{R} \citep{R},
also provide collections of various types of trees.
Within \proglang{R} the list of prominent packages includes
\pkg{rpart} \citep[implementing CART]{rpart},
\pkg{RWeka} \citep[containing interfaces to J4.8, M5', LMT from \pkg{Weka}]{RWeka}, and
\pkg{party} \citep[implementing CTree and MOB]{party} among
many others. See the CRAN task view ``Machine Learning''
\citep{ctv} for an overview.

All of these algorithms and software implementations have to deal with
very similar challenges. However, due to the fragmentation of the
communities in which the corresponding research is published -- ranging
from statistics over machine learning to various applied fields -- many
discussions of the algorithms do not reuse established theoretical results
and terminology. Similarly, there is no common ``language'' for the
software implementations and different solutions are provided by
different packages (even within \proglang{R}) with relatively little
reuse of code.

The \pkg{partykit} aims at mitigating the latter issue
by providing a common unified infrastructure
for recursive partytioning in the \proglang{R} system for statistical
computing. In particular, \pkg{partykit}
provides tools for representing fitted trees along with printing,
plotting, and computing predictions. The design principles are:
\begin{itemize}
  \item One `agnostic' base class (\class{party}) which can encompass an extremely wide
        range of different types of trees.
  \item Subclasses for important types of trees, e.g., trees with constant
        fits (\class{constparty}) or with parametric models (\class{modelparty})
	in each terminal node (or leaf).
  \item Nodes are recursive objects, i.e., a node can contain child nodes.
  \item Keep the (learning) data out of the recursive node and split structure.
  \item Basic printing, plotting, and predicting for raw node structure.
  \item Customization via suitable panel or panel-generating functions.
  \item Coercion from existing object classes in \proglang{R} (\code{rpart}, \code{J48}, etc.) to the new class.
  \item Usage of simple/fast \proglang{S}3 classes and methods.
\end{itemize}
In addition to all of this generic infrastructure, two specific tree
algorithms are implemented in \pkg{partykit} as well: \fct{ctree}
for conditional inference trees \citep{Hothorn+Hornik+Zeileis:2006}
and \fct{mob} for model-based recursive partitioning \citep{Zeileis+Hothorn+Hornik:2008}.

@@

While originally an important goal for the development of \pkg{partykit} was to
provide infrastructure for the authors' own tree induction algorithms
CTree and MOB,
the design was very careful to separate as much functionality as possible into
more general classes that are useful for a far broader class of tree algorithms.
In particular, to be able to print/plot/predict different trees in a unified way,
there are so-called coercion functions for transforming trees learned in other
software packages (inside and outside of \proglang{R}) into the classes provided
by \pkg{partykit}. Specifically, tree objects learned by \fct{rpart} from the
package of the same name (\citealp{rpart}, implementing CART, \citealp{Breiman+Friedman+Olshen:1984})
and by \fct{J48} from \pkg{RWeka} (\citealp{RWeka}, interfacing \pkg{Weka}'s J4.8
algorithm for C4.5, \citealp{Quinlan:1993}) can be coerced by \fct{as.party}
to the same object class \class{constparty}. This is a general class that can
in principle represent all the major classical tree algorithms with constant
fits in the terminal nodes. Also, the same class is employed for conditional
inference trees (CTree) that can be learned with the \fct{ctree} function
directly within \pkg{partykit} or evolutionary trees from package \pkg{evtree}
\citep{evtree}.

@@

All of these different trees (see Table~\ref{tab:overview} for an overview)
use the same infrastructure at the core but possibly with different options
enabled. In all cases, the functions \fct{print}, \fct{plot}, and \fct{predict}
can be used to create textual and graphical displays of the tree and for
computing predictions on new data, respectively. As an example for the
visualizations, Figure~\ref{fig:ttnc} shows two different trees fitted
to the well-known data on survival of passengers on the ill-fated maiden
voyage of the RMS~Titanic: The left panel shows a CART tree with constant fits
learned by \pkg{rpart} and converted to \pkg{partykit}. The right panel
shows a MOB tree learned within \pkg{partykit} with a logistic regression
model for treatment effects in the terminal nodes. Additionally, the
are further utility functions, e.g.,
\fct{nodeapply} can be employed to access further information
stored in the nodes of a tree and \fct{nodeprune} can prune selected nodes.

@@

especially for developers who either want to implement new tree algorithms
or represent trees learning in other systems in \proglang{R} to facilitate
predictions, comparisons, etc.

partysplit
partynode
party

constparty
simpleparty
modelparty

extractor functions

\section{Conclusions}

run-time, memory requirements, benchmarks?

@@

We are thankful to the organizers and participants of the ``Workshop on Classification and Regression Trees'' 2014,
sponsored by the Institute for Mathematical Sciences of the National University of Singapore,
for helpful feedback and stimulating discussions.

\newpage

>>>>

\begin{keywords}
recursive partitioning, regression trees, classification trees, statistical learning, \proglang{R}.
\end{keywords}

@@

In the more than fifty years since \cite{Morgan+Sonquist:1963}
published their seminal paper on ``automatic interaction detection'',
a wide range of methods has been suggested that is usually
termed ``recursive partitioning'' or ``decision trees'' or
``tree(-structured) models'' etc. The particularly influential 
algorithms include CART \citep[classification and regression trees,][]{Breiman+Friedman+Olshen:1984},
C4.5 \citep{Quinlan:1993}, QUEST/GUIDE \citep{Loh+Shih:1997,Loh:2002},
and CTree \citep{Hothorn+Hornik+Zeileis:2006} among many others
\citep[see][for a recent overview]{Loh:2014}. Reflecting the heterogeneity
of conceptual algorithms, a wide range of computational implementations
in various software systems emerged: Typically the original authors of an
algorithm also provide accompanying software but many software systems,
including \pkg{Weka} \citep{Witten+Frank:2005} or \proglang{R} \citep{R},
also provide collections of various types of trees.
Within \proglang{R} the list of prominent packages includes
\pkg{rpart} \citep[implementing CART]{rpart},
\pkg{RWeka} \citep[with interfaces to J4.8, M5', LMT from \pkg{Weka}]{RWeka}, and
\pkg{party} \citep[implementing CTree and MOB]{party} among
many others. See the CRAN task view ``Machine Learning''
\citep{ctv} for an overview.

All of these algorithms and software implementations have to deal with
similar challenges. However, due to the fragmentation of the
communities in which they are published -- ranging
from statistics over machine learning to various applied fields -- many
discussions of the algorithms do not reuse established theoretical results
and terminology. Similarly, there is no common ``language'' for the
software implementations and different solutions are provided by
different packages (even within \proglang{R}) with relatively little
reuse of code.
%
The \pkg{partykit} aims at mitigating the latter issue
by providing a common unified infrastructure
for recursive partytioning in the \proglang{R} system for statistical
computing. In particular, \pkg{partykit}
provides tools for representing, printing, plotting trees and
computing predictions. The design principles are:

\begin{itemize}
  \item One `agnostic' base class (\class{party}) encompassing a very wide
        range of different tree types.
  \item Subclasses for important types of trees, e.g., trees with constant
        fits (\class{constparty}) or with parametric models (\class{modelparty})
	in each terminal node (or leaf).
  \item Nodes are recursive objects, i.e., a node can contain child nodes.
  \item Keep the (learning) data out of the recursive node and split structure.
  \item Basic printing, plotting, and predicting for raw node structure.
  \item Customization via suitable panel or panel-generating functions.
  \item Coercion from existing object classes in \proglang{R} (\code{rpart}, \code{J48}, etc.) to the new class.
  \item Usage of simple/fast \proglang{S}3 classes and methods.
\end{itemize}
In addition to all of this generic infrastructure, two specific tree
algorithms are implemented in \pkg{partykit} as well: \fct{ctree}
for conditional inference trees \citep{Hothorn+Hornik+Zeileis:2006}
and \fct{mob} for model-based recursive partitioning \citep{Zeileis+Hothorn+Hornik:2008}.

@@

While originally an important goal for the development of \pkg{partykit} was to
provide infrastructure for the authors' own tree induction algorithms
CTree and MOB,
the design was very careful to separate as much functionality as possible into
more general classes that are useful for a far broader class of trees.
In particular, to be able to print/plot/predict different trees in a unified way,
there are so-called coercion functions for transforming trees learned in other
software packages (inside and outside of \proglang{R}) to the classes provided
by \pkg{partykit}. Specifically, tree objects learned by \fct{rpart}
(\citealp{rpart}, implementing CART, \citealp{Breiman+Friedman+Olshen:1984})
and by \fct{J48} from \pkg{RWeka} (\citealp{RWeka}, interfacing \pkg{Weka}'s J4.8
algorithm for C4.5, \citealp{Quinlan:1993}) can be coerced by \fct{as.party}
to the same object class \class{constparty}. This is a general class that can
in principle represent all the major classical tree algorithms with constant
fits in the terminal nodes. Also, the same class is employed for conditional
inference trees (CTree) that can be learned with the \fct{ctree} function
directly within \pkg{partykit} or evolutionary trees from package \pkg{evtree}
\citep{evtree}.

@@

All of these different trees (see Table~\ref{tab:overview} for an overview)
use the same infrastructure at the core but possibly with different options
enabled. In all cases, the functions \fct{print}, \fct{plot}, and \fct{predict}
can be used to create textual and graphical displays of the tree and for
computing predictions on new data, respectively. As an example for the
visualizations, Figure~\ref{fig:ttnc} shows two different trees fitted
to the well-known data on survival of passengers on the ill-fated maiden
voyage of the RMS~Titanic: The left panel shows a CART tree with constant fits
learned by \pkg{rpart} and converted to \pkg{partykit}. The right panel
shows a MOB tree learned with \pkg{partykit} with a logistic regression
for treatment effects in the terminal nodes. Additionally, the
are further utility functions, e.g.,
\fct{nodeapply} can be employed to access further information
stored in the nodes of a tree and \fct{nodeprune} can prune selected nodes.

@@

The unified infrastructure at the core of \pkg{partykit} is especially
appealing for developers who either want to implement new tree algorithms
or represent trees learned in other systems.

Here, we briefly outline
the most important classes. More details can be found in the vignettes.
\begin{description}
  \item[\class{partysplit}:] Split specification with an integer ID for the
    splitting variable, breakpoint(s), and indexes for the kids.
  \item[\class{partynode}:] Node specification with an integer ID, a
    \class{partysplit}, and a list of kids (if any) that are \class{partynode}
    objects again.
  \item[\class{party}:] Tree specification with a \class{partynode} tree
    along with a \class{data.frame} (which optionally may be empty),
    potentially plus further information about fitted values and so-called
    \class{terms} that allow to preprocess new data for predictions.
\end{description}
All classes have an additional slot for storing arbitrary information at
any level of the tree. This is exploited by \class{constparty}, \class{simpleparty},
and \class{modelparty} which store the observed response, point predictions,
and fitted parametrics models, respectively.
% Extractor functions for the various components of splits, nodes, and trees are provided.

\section{Discussion and Outlook}

Package \pkg{partykit} provides a toolkit for trees in \proglang{R} that gives
emphasis to flexibility and extensibility. The infrastructure is easily
accessible and written in plain \proglang{R}. Nevertheless speed is not a critical
concern because (a)~lean data structures are employed and (b)~the time-intensive
part in tree computations is typically inferring the tree rather than
representing/printing/plotting it. Hence, the two functions for inferring trees
either have computationally intensive parts in \proglang{C} (\fct{ctree})
or build on \proglang{R}'s fitting functions (\fct{mob}). For benchmarks of
their predictive performance, see their original references.
Regarding memory requirements, the \class{party} trees are very lean but can
optionally be used to store any information. The recursive node structure
itself is extremely lightweight and optionally zero rows from the learning data
are sufficient, which future versions will exploit for storing forests.

@@

We are thankful to the organizers and participants of the
``Workshop on Classification and Regression Trees'' (March 2014),
sponsored by the Institute for Mathematical Sciences of the
National University of Singapore, for helpful feedback and
stimulating discussions.
